Autopolyploid establishment through polygenic adaptation

### 2025-01-14 doc/paper/paper.tex
update

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics (\textit{sensu}
\cite{barton2017}) for the inheritance of an additive quantitative trait in a
mixed-ploidy population consisting of diploid, triploid and autotetraploid
individuals producing haploid and diploid gametes.
We implement efficient simulation methods and use these to study the
quantitative genetics of mixed-ploidy populations and the establishment of
autotetraploids in a new habitat.
We show that for rare migration from the source population, autotetraploids are
more likely to found a successful population than diploids under a very
broad range of conditions, but that this is unlikely to sufficiently
counter the scarcity of tetraploid founders when the source is
predominantly diploid.

\end{abstract}

@@

In this study, we worked out the infinitesimal model for an additive polygenic
trait in autotetraploids and mixed-ploidy populations.
We used the mixed-ploidy infinitesimal model to study the establishment of
tetraploids in a marginal habitat by means of individual-based simulations. 
Assuming the trait to be under directional selection in the marginal habitat,
and migration of maladapted individuals from a predominantly diploid source,
we sought to determine the conditions under which tetraploids are more likely
to establish a stable population.
Importantly, we assume no intrinsic advantage of polyploids in the marginal
habitat, i.e. the expected fitness of a migrant individual is the same
regardless of the ploidy level.
Differences in establishment probabilities are hence caused solely by aspects
of autopolyploid genetics and the barrier to gene flow between subpopulations
of different ploidy levels.

@@

It would be very interesting to experimentally assess whether typical
quantitative traits in mixed-ploidy populations follow approximately the
mixed-ploidy infinitesimal model we outlined in this paper, and to estimate the
relevant model parameters (in particular the $\beta_3$ and $\beta_4$ values).

>>>>

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics (\textit{sensu}
    \cite{barton2017}) for the inheritance of an additive quantitative trait in
    a mixed-ploidy population consisting of diploid, triploid and
    autotetraploid individuals producing haploid and diploid gametes.  We
    implement efficient simulation methods and use these to study the
    quantitative genetics of mixed-ploidy populations and the establishment of
    autotetraploids in a new habitat.  We show that for rare migration from the
    source population, autotetraploids are more likely to found a successful
    population than diploids under a very broad range of conditions, but that
    this is unlikely to sufficiently counter the scarcity of tetraploid
    founders when the source is predominantly diploid.  We assess in more
    detail how minority cytotype exclusion interacts with migration load in the
    establishment process and evaluate the impact of additional sources of
    prezygotic isolation, specifically selfing and assortative mating, on the
    relative establishment probabilities of the different cytotypes.
    In the discussion we consider how inbreeding depression may impact our
    findings.
\end{abstract}

@@

In this study, we worked out the infinitesimal model for an additive polygenic
trait in autotetraploids and mixed-ploidy populations.
It would be very interesting to experimentally assess whether typical
quantitative traits in mixed-ploidy populations follow approximately the
mixed-ploidy infinitesimal model we outlined in this paper, and to estimate the
relevant model parameters (in particular the $\beta_3$ and $\beta_4$ values).

We used the mixed-ploidy infinitesimal model to study the establishment of
tetraploids in a marginal habitat by means of individual-based simulations. 
Assuming the trait to be under directional selection in the marginal habitat,
and migration of maladapted individuals from a predominantly diploid source,
we sought to determine the conditions under which tetraploids are more likely
to establish a stable population.
Importantly, we assume no intrinsic advantage of polyploids in the marginal
habitat, i.e. the expected fitness of a migrant individual is the same
regardless of the ploidy level.
Differences in establishment probabilities are hence caused solely by aspects
of autopolyploid genetics and the barrier to gene flow between subpopulations
of different ploidy levels.

@@
